The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-10-06 trunk/vcd/paper/Strucplot/JSS.Rnw
second bunch of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@582 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of `graphical
appearance control' functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is given in Appendix \ref{sec:code}). 
Figure~\ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. Figure~\ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The $\chi^2$ test 
of independence rejects the null hypothesis: possibly, because
the tendency of people  to have extramarital
sex when they had premarital sex 
is particularly marked among married people?
Finally, Figure~\ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival of the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
@
\caption{Mosaic plot for the \data{PreSex} data.}
\label{fig:presex}
\end{center}
\end{figure}

@@

The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

The strucplot framework is highly modularized: Figure~\ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of grapcon functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure~\ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\textit{group\_foo}()},  
where (\code{\textit{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\textit{foo}()}, 
\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\textit{foo}()} implement the core functionality, 
creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous gender variable (\code{Sex}). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

@@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
the default method for \codefun{doubledecker} is simply a wrapper for 
\codefun{strucplot}, setting the right defaults.
Most default settings such as colors, spacing, and labeling are
specified via the parameters and passed through to
\codefun{strucplot}. The additional code just handles the dependent
variable information, and in particular permutes the table to have the dependent
variable as the last dimension as required for the doubledecker plot.
Figure~\ref{fig:titanic} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
sex, given class. It is created by:

@@

So far, we have visualized either full or collapsed tables, as
suggested by the analysis task at hand. Subtables can be selected in a
similar way as for objects of class \class{table} using
indexing. Note, however, that subsetting of \class{structable} objects 
is more restrictive because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
subsetting is only possible ``outside-in'', that is, indexing operates
on blocks defined by the variable levels. In the following, we use the
Titanic data again, this time collapsed over `Survived' to investigate the
structure of crew and passengers (and having the \code{Child} and
\code{Age} labels of the \code{Age} variable swapped for optical clarity):

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables1>>
<<Variables2>>
<<Variables3>>
@
\caption{Two mosaic displays put side-by-side, visualizing the 
distribution of class and age, given gender. The marginal
distribution of `Sex' cannot be seen.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{Titanic} data,
  again visualizing the distribution of age and class, given
  gender, using separate mosaic displays like the `manual' plot 
  in Figure~\ref{fig:parttable}.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}
%\noindent The \code{labeling\_args} argument modifies the labels'
%appearance: here, to be left-aligned and unclipped
%(see Section \ref{sec:labeling}).

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning2>>
@
\caption{Mosaic plot again visualizing the 
distribution of class and age, given gender, this time using a
single mosaic plot. In contrast to Figures~\ref{fig:parttable} and \ref{fig:cotabplot}, 
this plot also visualizes the marginal distribution of \code{Sex}.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
frequencies under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

@@

\begin{enumerate}
\item Precomputing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
 \item Providing a grapcon \emph{generating} function (`grapcon generator') 
  taking parameters and returning a function as described in alternative~2.
\end{enumerate}

@@

The first method is precomputing the graphical parameters ``by
hand''. We will use `light blue' color for positive and `light salmon'
color for negative residuals (see Figure~\ref{fig:binary}):

<<shading1,results=verbatim>>=
expected <- independence_table(ucb)
(residuals <- (ucb - expected)  / sqrt(expected))

(shading1_obj <- ifelse(residuals > 0, "lightblue", "lightsalmon"))

@@

For implementing alternative 2, we need to create a `shading function'
that computes \class{gpar} objects from
residuals. For that, we can just reuse the code from the previous step:

<<shading2>>=
shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
@

@@

Our \codefun{shading2\_fun} function might be useful, 
but can still be improved: the hard-wired colors should be
customizable. We cannot
simply extend the argument list to include, e.g., 
a \code{fill = c("lightblue", "lightsalmon")} argument because
\codefun{strucplot} will neither know how to handle it, nor let us
change the defaults. In fact, the interface of shading functions is fixed,
they are expected to take exactly one argument: a table of residuals. 
This is where generating functions (alternative 3) come
into play.

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in \proglang{R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

<<shading3>>=
shading3a_fun <- function(col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
@

\noindent In the call to \codefun{mosaic}, using the new
\codefun{shading3a\_fun} function, we can now simply change the colors:

<<shading4>>=
mosaic(ucb, gp = shading3a_fun(c("red","blue")))
@

@@

<<shading4>>=
shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
    df = NULL, col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
class(shading3b_fun) <- "grapcon_generator"
@

In some sense, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:

@@

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using HSV space, and
then briefly explain the differences to HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
Color palettes in HCL space are trickier
to choose, but preferable to palettes derived 
from HSV space from a perceptual point of view.

In HSV space, colors are specified in three dimensions: Hue,
Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
These three dimensions are used by \codefun{shading\_hsv} to visualize information 
about the residuals and the underlying independence model. The hue 
indicates the residuals' sign: by default, blue for positive, and red
for negative residuals. The saturation of a residual 
is set according to its size: high saturation for large, and low
saturation for small residuals. 
Finally, the overall lightness is used to indicate the significance of
a test statistic: light colors for significant, 
and dark colors for non-significant results.

As an example, we will visualize the association of hair and eye
color in the \data{HairEyeColor} data set (see Figure~\ref{fig:haireye})

<<haireye>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, gp = shading_hsv)
@

\begin{figure}[h]
\begin{center}
<<haireyefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<haireye>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent Large positive residuals (greater than $4$) can be found for
brown eyes/black hair and blue eyes/blond hair, and are colored in saturated blue. 
On the other hand, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
are shaded in white.
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure~\ref{fig:interpolate}):

<<interpolate>>=
mosaic(haireye, gp = shading_hsv, gp_args = list(interpolate = 1:4))
@

\begin{figure}[h]
\begin{center}
<<interpolatefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<interpolate>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---four cut-off points.}
\label{fig:interpolate}
\end{center}
\end{figure}

@@

\noindent Note that this \codefun{ipol} 
function maps residuals greater than 4 to a saturation
level of 1. However, the resulting plot (Figure~\ref{fig:continuous}) is 
deceiving:

<<continuous2>>=
mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
labeling_args = list(abbreviate = c(Sex = TRUE)))
@

\begin{figure}[p]
\begin{center}
<<continuousfig,fig=TRUE,echo=FALSE,height=7>>=
<<continuous2>>
@
\caption{The \data{HairEyeColor} data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

@@

As an example, consider the \data{Titanic} data, consisting
of 4 categorical variables: survival (`survived'), gender (i.e., `sex'), age, and
crew. By default, the variable names and levels are plotted `around' the
plot in a counter-clockwise way (see Figure~\ref{fig:defaults}):

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% an abstract and keywords
\Abstract{
This paper describes the ``strucplot'' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of ``graphical
appearance control'' functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://statmath.wu-wien.ac.at/~hornik/}\\

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is given in Appendix \ref{sec:code}). 
Figure~\ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. The dark
blue cell indicates that the rate of treated patients showing marked improvement
is significant at the 1\% level. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. Figure~\ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The $\chi^2$ test 
of independence, based on the permutation distribution,  
rejects the null hypothesis: possibly, because
the tendency of people to have extramarital
sex when they had premarital sex 
is particularly marked among married people? The rate of such women and men
ist significant at the 0.01 and 0.1 level, respectively.
Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
data using a double-decker plot. Here, a binary response (survival of the disaster)
is to be explained by other factors (class, gender, and age). The gray
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
presextest <- coindep_test(PreSex, margin = c(1,4), 
                           indepfun = function(x) sum(x^2), n = 5000)
mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
       gp_args = list(p.value = presextest$p.value))
@
\caption{Mosaic plot for the \data{PreSex} data.}
\label{fig:presex}
\end{center}
\end{figure}

@@

The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
``flat'' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

The strucplot framework is highly modularized: Figure~\ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of grapcon functions. They are created by generating functions
(\emph{grapcon generators}), allowing
flexible parameterization and extensibility (Figure~\ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\textit{group\_foo}()},  
where (\code{\textit{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\textit{foo}()}, 
\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\textit{foo}()} implement the core functionality, 
creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous gender variable (\code{Sex}). The
``flattened'' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

@@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
the default method for \codefun{doubledecker} is simply a wrapper for 
\codefun{strucplot}, setting the right defaults.
Most default settings such as colors, spacing, and labeling are
specified via the parameters and passed through to
\codefun{strucplot}. The additional code just handles the dependent
variable information, and in particular permutes the table to have the dependent
variable as the last dimension as required for the doubledecker plot.
Figure~\ref{fig:titanic} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (``survived'') by age, given
sex, given class. It is created by:

@@

So far, we have visualized either full or collapsed tables, as
suggested by the analysis task at hand. Subtables can be selected in a
similar way as for objects of class \class{table} using
indexing. Note, however, that subsetting of \class{structable} objects 
is more restrictive because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
subsetting is only possible ``outside-in'', that is, indexing operates
on blocks defined by the variable levels. In the following, we use the
Titanic data again, this time collapsed over \code{Survived} to investigate the
structure of crew and passengers (and having the \code{Child} and
\code{Age} labels of the \code{Age} variable swapped for optical clarity):

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables1>>
<<Variables2>>
<<Variables3>>
@
\caption{Two mosaic displays put side-by-side, visualizing the 
distribution of class and age, given gender. The marginal
distribution of gender cannot be seen.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{Titanic} data,
  again visualizing the distribution of age and class, given
  gender, using separate mosaic displays like the ``manual'' plot 
  in Figure~\ref{fig:parttable}.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}
%\noindent The \code{labeling\_args} argument modifies the labels'
%appearance: here, to be left-aligned and unclipped
%(see Section \ref{sec:labeling}).

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning2>>
@
\caption{Mosaic plot again visualizing the 
distribution of class and age, given gender, this time using a
single mosaic plot. In contrast to Figures~\ref{fig:parttable} and \ref{fig:cotabplot}, 
this plot also visualizes the marginal distribution of gender.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
frequencies under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading 
\citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005},
even allows to identify the cells that cause the rejection of the null hypothesis.

@@

\begin{enumerate}
\item Precomputing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
 \item Providing a grapcon generator taking parameters and 
   returning a function as described in alternative~2.
\end{enumerate}

@@

The first method is precomputing the graphical parameters ``by
hand''. We will use \code{royalblue4} color for positive and \code{mediumorchid4}
color for negative residuals (see Figure~\ref{fig:binary}):

<<shading1,results=verbatim>>=
expected <- independence_table(ucb)
(x <- (ucb - expected)  / sqrt(expected))

(shading1_obj <- ifelse(x > 0, "royalblue4", "mediumorchid4"))

@@

For implementing alternative 2, we need to create a ``shading function''
that computes \class{gpar} objects from
residuals. For that, we can just reuse the code from the previous step:

<<shading2>>=
shading2_fun <- function(x) gpar(fill = ifelse(x > 0, "royalblue4", "mediumorchid4"))
@

@@

Our \codefun{shading2\_fun} function might be useful, 
but can still be improved: the hard-wired colors should be
customizable. We cannot
simply extend the argument list to include, e.g., 
a \code{fill = c("royalblue4", "mediumorchid4")} argument because
\codefun{strucplot} will neither know how to handle it, nor let us
change the defaults. In fact, the interface of shading functions is fixed,
they are expected to take exactly one argument: a table of residuals. 
This is where generating functions (alternative 3) come
into play.

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in \proglang{R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (``creates'') a 
``parameterized'' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

<<shading3>>=
shading3a_fun <- function(col = c("royalblue4", "mediumorchid4")) {
  col <- rep(col, length.out = 2)
  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
}
@

\noindent The first statement just makes sure that exactly two colors
are specified. In the call to \codefun{mosaic}, using the new
\codefun{shading3a\_fun} function, we can now simply change the colors:

<<shading4>>=
mosaic(ucb, gp = shading3a_fun(c("royalblue4","mediumorchid4")))
@

@@

<<shading4>>=
shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
    df = NULL, col = c("royalblue4", "mediumorchid4")) {
  col <- rep(col, length.out = 2)
  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
}
class(shading3b_fun) <- "grapcon_generator"
@

Note that in this simple binary shading example, the first four
parameters are not used. In some sense, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:

@@

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using HSV space, and
then briefly explain the differences to HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
Color palettes in HCL space are 
preferable to palettes derived 
from HSV space from a perceptual point of view. Functions for creating
palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the package.

In HSV space, colors are specified in three dimensions: Hue,
Saturation (``colorfulness''), and Value (``lightness'', amount of gray). 
These three dimensions are used by \codefun{shading\_hsv} to visualize information 
about the residuals and the underlying independence model. The hue 
indicates the residuals' sign: by default, blue for positive, and red
for negative residuals. The saturation of a residual 
is set according to its size: high saturation for large, and low
saturation for small residuals. 
Finally, the overall lightness is used to indicate the significance of
a test statistic: light colors for significant, 
and dark colors for non-significant results.

As an example, we will visualize the association of hair and eye
color in the \data{HairEyeColor} data set (see
Figure~\ref{fig:haireye}, top):

<<haireye1>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, gp = shading_hsv)
@

\noindent As introduced before, the default shading scheme is not
\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
perceptual characteristics of HCL color space.
The following example again illustrates the \data{HairEyeColor} data, 
this time with HCL colors:

<<haireye2>>=
mosaic(haireye, gp = shading_hcl)
@ 
<<haireye3>>=
mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
@

\noindent In Figure~\ref{fig:haireye}, the plot
in the middle depicts the default palette, 
and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{s}), and
Luminance (\code{v}).

\setkeys{Gin}{width=0.7\textwidth}

\begin{figure}[htbp]
\begin{center}
<<haireyefig1,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig2,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig3,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
@
\caption{Three mosaic plots for the \data{HairEyeColor} data using, from top to bottom:
the default HSV color palette, the default HCL color palette, 
and a custom HCL color palette.}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent Large positive residuals (greater than $4$) can be found for
brown eyes/black hair and blue eyes/blond hair, and are colored in deep blue. 
On the other hand, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
are shaded in white (gray for HCL-shading).
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure~\ref{fig:interpolatecontinuous}):

<<interpolate>>=
mosaic(haireye, shade = TRUE, gp_args = list(interpolate = 1:4))
@

@@

\noindent Note that this \codefun{ipol} 
function maps residuals greater than 4 to a saturation
level of 1. However, the resulting plot
(Figure~\ref{fig:interpolatecontinuous}, right) is 
deceiving:

<<continuous2>>=
mosaic(haireye, shade = TRUE, gp_args = list(interpolate = ipol),
labeling_args = list(abbreviate = c(Sex = TRUE)))
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<interpolatefig,fig=TRUE,echo=FALSE,width=12,height=5>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))
pushViewport(viewport(layout.pos.col = 1))
mosaic(haireye, gp_args = list(interpolate = 1:4), margin = c(right = 1), keep_aspect_ratio= FALSE,newpage = FALSE,legend_width=5.5,shade = TRUE)
popViewport(1)
pushViewport(viewport(layout.pos.col = 2))
mosaic(haireye, gp_args = list(interpolate = ipol), margin = c(left=3,right = 1), keep_aspect_ratio = FALSE, newpage = FALSE, shade = TRUE)
popViewport(2)
@
\caption{\label{fig:interpolatecontinuous}The \data{HairEyeColor} data---left: with 4 cut-off points, right: with continuous shading.}

\end{center}
\end{figure} 
\setkeys{Gin}{width=0.7\textwidth}

@@

As an example, again consider the \data{Titanic} data:
by default, the variable names and levels are plotted ``around'' the
plot in a counter-clockwise way (see Figure~\ref{fig:labels1}, top left):
